---
title: "HW#2e6-main"
output: html_document
---

```{r setup, include=FALSE, ref.label="sources"}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 6. Clustering: Cluster the January and July precipitation over Colorado 
January and July data, together with functions and libraries are loaded at the end of the document.

### (i) Identify the best number of clusters (Kbest) using lat, long, elev

To identify the best K we use the WSS distances of the scaled data.
```{r}
datajan=scale(jan) %>% data.frame(.)
datajul=scale(jul) %>% data.frame(.)

par(mfrow=c(1,2))
wssplot(datajan,nc=30) # Plots With-In-Sum-Of-Squares (elbow curve) # See source chunk
wssplot(datajul,nc=30) # Plots With-In-Sum-Of-Squares (elbow curve) # See source chunk
```

### (ii) Cluster the data into Kbest clusters and display them (using lat, long, elev)

There is no clear elbow in the data, but we can assume differences are relatively small once k is around 7.

```{r}
size=7

plotkmeans(jan,size) # Plots clusters spatially, function defined at the end
plotkmeans(jul,size) # jul

```

### (iii) Identify the best number of clusters (Kbest) using lat, long

```{r}
datajan=scale(jan[!colnames(jan) %in% "Elev"]) %>% data.frame(.)
datajul=scale(jul[!colnames(jul) %in% "Elev"]) %>% data.frame(.)

par(mfrow=c(1,2))
wssplot(datajan,nc=30) # Plots With-In-Sum-Of-Squares (elbow curve) # See source chunk
wssplot(datajul,nc=30) # Plots With-In-Sum-Of-Squares (elbow curve) # See source chunk
```

### (iv) Cluster the data into Kbest clusters and display them (using lat, long)

Again, there is no clear elbow in the data, but we can assume differences are relatively small once k is around 10. We will choose 7 to compare them better with the previous clusterings.

```{r}
size=7

plotkmeans(jan,size) # Plots clusters spatially, function defined at the end
plotkmeans(jul,size) # jul

```

The maps are similar to the previous ones, showing that *Elevation* is not relevant.

#### Aux code to call libraries, own functions and variables

```{r sources, results=FALSE}
##
## Source Libraries
##

libr=c("stats","magrittr","maps","colorRamps")
options(warn=-1)
suppressPackageStartupMessages(lapply(libr, require,character.only = TRUE))


##
## Source data
##

### January and July precipitation
jan=read.table("http://cires1.colorado.edu/~aslater/CVEN_6833/colo_monthly_precip_01.dat")
colnames(jan)<-c("Lat","Long","Elev","Prec")

jul=read.table("http://cires1.colorado.edu/~aslater/CVEN_6833/colo_monthly_precip_07.dat")
colnames(jul)<-c("Lat","Long","Elev","Prec")

data=list(jan,jul) #Set up all data
datan<-list("jan","jul")
for(i in 1:length(data)){
  datanames<-colnames(data[[i]])          # temp. Headings
  data[[i]]<-data[[i]][c("Prec",datanames[!datanames %in% "Prec"])] #Predictor first
  data[[i]]$Prec<-ifelse(data[[i]]$Prec<=0,NA,data[[i]]$Prec) #Eliminate zero-neg values
  data[[i]]<-data[[i]][complete.cases(data[[i]]), ] # Clean data
}
for(i in 1:length(datan)){assign(as.character(datan[[i]]),data[[i]])}

##
## Source functions
## 

### Plot With-In-Sum-Of-Squares  #
wssplot <- function(data, nc=15, seed=1234){
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares",main = substitute(data))}

### Plot spatial clustering data  #
plotkmeans = function(data,size){
  clust=kmeans(scale(data),size) # clustering k=size
  print("Number of elements of each cluster: ")
  print(clust$size)          # N at each cluster

  plot(data$Long, data$Lat, col = primary.colors()[clust$cluster], xlab = "Longitude", ylab = "Latitude", main = paste("Clusters for",substitute(data) ,"precipitation of Colorado (1980-2002) - K =", size), pch = 16, asp = 1)
  map('county', 'colorado',col = "grey", lwd = 2, add=TRUE)
  }
```
