---
title: "HW#2e3_main"
output: html_document
---

```{r setup, include=FALSE, ref.label="sources"}
knitr::opts_chunk$set(echo = TRUE)
```
## Exercise 3. Principal Component Analysis 

### (i) Perform a PCA on the winter global SST anomalies
Source libraries, data and functions. Import and setup SST data
```{r kaplan}
setwd("C:/Users/alexb/Google Drive/CVEN 6833 ADAT/zz Homeworks/HW#2/CVEN6833-HW-2")
set.seed(1) # Set seed to repodruce results
# Libraries, functions (and variables) are sourced from the last chunk ("sources") at the end of the page.

require(magrittr)

nrows=72 # data import setup
ncols=36
nyrs = 118  # Nov-Mar avg 1901 - 2018

xgrid=seq(27.5,382.5,by=5) #Generate grid (-180,180) format
xgrid[xgrid>180]=xgrid[xgrid>180]-360
ygrid=seq(-87.5,87.5,by=5)
xygrid=expand.grid(x=xgrid,y=ygrid) %>% .[c("y","x")] # Lat - Long grid..

kaplan = # SST data row per NDJFM season format
  readBin("http://civil.colorado.edu/%7Ebalajir/CVEN6833/HWs/HW-2-2018/Kaplan-SST-NDJFM1900-NDJFM2018.r4",what="numeric", n=( nrows * ncols * nyrs), size=4,endian="swap") %>% # Read Kaplan SST data..
  array(data = ., dim=c( nrows*ncols, nyrs ) ) %>% t(.) # 2D

index1=kaplan[1,]<10 & kaplan[1,] != "NaN"
sstdata=kaplan[,index1] # Drop missing data
```

The PCA is performed using the SST data. First we calculate the variance matrix, SVD decomposition, and Eigenvalues.

```{r PCA_kaplan}
zsvd = sstdata %>% scale(.) %>% var(.) %>% svd(.) # scale # get variance # Eigen decomposition
pcs=t(t(zsvd$u) %*% t(scale(sstdata))) #Principal Components...
lambdas=(zsvd$d/sum(zsvd$d)) #Eigen Values.. - fraction variance explained 
```

#### Plot of variance explained by each principal component

```{r }
plot(1:40, lambdas[1:40], type="l", xlab="Modes", ylab="Frac. Var. explained")
points(1:40, lambdas[1:40], col="red")
```

#### First four spatial components or Eigen Vector patterns

```{r}
xlong = sort(xgrid,index.return=TRUE)
ylat = sort(ygrid,index.return=TRUE)

for(i in 1:4){
  zfull = rep(NA,length(xygrid[,1]))
  zfull[index1]=zsvd$u[,i]
  zmat = matrix(zfull,nrow=nrows,ncol=ncols)
  zmat = zmat[xlong[["ix"]],] %>% .[,ylat[["ix"]]] # match SST values with sorted rows,columns

  image.plot(xlong[["x"]],ylat[["x"]],zmat,ylim=range(-40,70),main=paste("Spatial Component (eigenvector) number",i))
  contour(xlong[["x"]],ylat[["x"]],zmat,ylim=range(-40,70),add=TRUE,nlev=6,lwd=2)
  world(fill=TRUE, col="darkgray",border="gray",add=TRUE)
}
```

#### First four Principal Components (time)

```{r fig.width=16}
par(mfrow=c(2,2))
nino4=scan("http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/.NINO4/T/(Nov%201899)/(Mar%202018)/RANGE/T/(Nov-Mar)/seasonalAverage/data.ch") # ENSO data

for(i in 1:4){
  plot(seq(from=1901,length.out=nyrs), scale(pcs[,i]),type="b",xlab="Year",ylab=paste("PC",i),main=paste("Temporal Component (eigenvalue) number",i))
  if(i==2){lines(1901:2018, nino4, col="red") 
    legend(x=1,y=1,legend=c("PC","ENSO"))}
}
```
### (ii) Rotate the first 6 PCs and plot the leading 4 spatial and temporal modes of variability
There are three ways to do it, see: <https://stats.stackexchange.com/questions/59213/how-to-compute-varimax-rotated-principal-components-in-r>
We will use the varimax function and calculate the scores via rotation matrix
```{r}
zrot = varimax(zsvd$u[,1:6],normalize=FALSE)
for(i in 1:4){
  zfull = rep(NA,length(xygrid[,1]))
  zfull[index1]=zrot$loadings[,i]
  zmat = matrix(zfull,nrow=nrows,ncol=ncols)
  zmat = zmat[xlong[["ix"]],] %>% .[,ylat[["ix"]]] # match SST values with sorted rows,columns

  image.plot(xlong[["x"]],ylat[["x"]],zmat,ylim=range(-40,70),main=paste("Rotated spatial component (eigenvector) number",i))
  contour(xlong[["x"]],ylat[["x"]],zmat,ylim=range(-40,70),add=TRUE,nlev=6,lwd=2)
  world(fill=TRUE, col="darkgray",border="gray",add=TRUE)
  }
```
The use of rotated PCA allows us to observe a more acute spatial distribution of the eigenvectors

The rotated temporal components are:

```{r fig.width=16}
par(mfrow=c(2,2))
rotpcs=t(t(zrot$loadings) %*% t(scale(sstdata)))
for(i in 1:4){
  plot(seq(from=1901,length.out=nyrs), scale(rotpcs[,i]),type="b",xlab="Year",ylab=paste("PC",i),main=paste("Rotated Temporal Component (eigenvector) number",i))
    if(i==2){lines(1901:2018, nino4, col="red") 
    legend(x=1,y=1,legend=c("PC","ENSO"))}
}
```

This temporal components seem similar to the originals.

### (iii) Perform PCA and rotated PCA on the western US winter precipitation (Nov1900 – Mar2013) 113
yrs
First we import grid location and precipitation data
```{r}
N = 113 # Time window
wuplocs = matrix(scan("http://civil.colorado.edu/%7Ebalajir/CVEN6833/HWs/HW-2-2018/WesternUS-coords.txt"), ncol=5,byrow=T)
nlocs = dim(wuplocs)[1]
xlats = wuplocs[,3]
xlongs = -wuplocs[,4]

winterprecip = matrix(scan("http://civil.colorado.edu/%7Ebalajir/CVEN6833/HWs/HW-2-2018/western-US-winterprecip-1901-2014.txt"),ncol=nlocs+1,byrow=T)
years = 1901:2013
winterprecip = winterprecip[,2:(nlocs+1)]		#first column is year
```

#### Do PCA on precipitation data

```{r}
zsvd = winterprecip %>% scale(.) %>% var(.) %>% svd(.) # Eigen decomposition
pcs=t(t(zsvd$u) %*% t(scale(winterprecip))) # Principal Components
lambdas=(zsvd$d/sum(zsvd$d)) # Eigenvalues
```

#### Variance explained by each principal component

```{r }
plot(1:25, lambdas[1:25], type="l", xlab="Modes", ylab="Frac. Var. explained")
points(1:25, lambdas[1:25], col="red")
```

#### First four spatial components or Eigen Vector patterns

```{r}
for(i in 1:4){
  quilt.plot(xlongs,xlats,zsvd$u[,i],xlim=range(-125,-100),main=paste("Spatial Component (eigenvector) number",i))
  maps::map("state",add=TRUE)
}
```

#### First four Principal Components (time)

```{r fig.width=16}
par(mfrow=c(2,2))
westernpcs=as.data.frame(matrix(, nrow = length(pcs[,1]), ncol = 4))
colnames(westernpcs)=c("PC1","PC2","PC3","PC4")
for(i in 1:4){
  westernpcs[,i]=scale(pcs[,i])
  plot(seq(from=1901,length.out=N),scale(pcs[,i]),type="b",xlab="Year",ylab=paste("PC",i),main=paste("Temporal Component (eigenvalue) number",i))
}
```

#### Rotation of the first 6 first PCs and plot of the leading 4 spatial and temporal modes of variability
```{r}
zrot = varimax(zsvd$u[,1:6],normalize=FALSE)
for(i in 1:4){
  quilt.plot(xlongs, xlats, zrot$loadings[,i],xlim=range(-125,-100),main=paste("Rotated spatial mode (Eigenvector) number",i))
  map("state",add=TRUE)
}
```

#### Plot of the first four rotated PC
A simple glm is added to show trend.
```{r fig.width=16}
par(mfrow=c(2,2))
rotpcs=t(t(zrot$loadings) %*% t(scale(winterprecip)))
for(i in 1:4){
  plot(seq(from=1901,length.out=N), scale(rotpcs[,i]),type="b",xlab="Year",ylab=paste("PC",i),main=paste("Rotated temporal Component (eigenvalue) number",i))
  if(i==1){
    abline(glm(rotpcs[,i]~seq(from = 1901,length.out = N)), col="blue", lwd=2) # trend
  }
  } 
```

### (iv) Perform a PCA on the Pacific Ocean domain and correlate the first four PCs of winter SSTs with that of the western US precipitation and show correlation maps.

To perform PCA on the Pacific Ocean grid only, we retrieve kaplan and trim grid to study area

```{r ref.label="kaplan", echo=FALSE}
```

```{r}
pacgrid=xygrid$y >= -20 & (xygrid$x >= 105 | xygrid$x <= -70) #grid limits
indexpac=index1 & pacgrid # locations on kaplan grid
xygrid1 = xygrid[indexpac,] # coordinates on kaplan grid

sstannavg = kaplan[,indexpac] # Pacific data 
```

Now we do PCA on data

```{r}
zsvd = sstannavg %>% scale(.) %>% var(.) %>% svd(.) # SVD decomposition
pcs=t(t(zsvd$u) %*% t(scale(sstannavg)))
lambdas=(zsvd$d/sum(zsvd$d))
```

#### Plot of variance explained by each principal component

```{r }
plot(1:25, lambdas[1:25], type="l", xlab="Modes", ylab="Frac. Var. explained")
points(1:25, lambdas[1:25], col="red")
```

#### First four spatial components or Eigen Vector patterns

```{r}
xgrid=seq(27.5,382.5,by=5) #Generate grid (27.5,382.5) format
xlong = sort(xgrid,index.return=TRUE)
ylat = sort(ygrid,index.return=TRUE)

for(i in 1:4){
  zfull = rep(NA,length(xygrid[,1]))
  zfull[indexpac]=zsvd$u[,i]
  zmat = matrix(zfull,nrow=nrows,ncol=ncols)
  zmat = zmat[xlong[["ix"]],] %>% .[,ylat[["ix"]]] # match SST values with sorted rows,columns

  image.plot(xlong[["x"]],ylat[["x"]],zmat,ylim=range(-40,70),main=paste("Spatial Component (eigenvector) number",i))
  contour(xlong[["x"]],ylat[["x"]],zmat,ylim=range(-40,70),add=TRUE,nlev=6,lwd=2)
  maps::map("world2",fill = TRUE,col="darkgray",border="gray",add=TRUE)
}
```

#### First four Principal Components (time)

```{r fig.width=16}
par(mfrow=c(2,2))
pacificpcs=as.data.frame(matrix(, nrow = length(pcs[,1]), ncol = 4))
colnames(pacificpcs)=c("PC1","PC2","PC3","PC4")
for(i in 1:4){
  pacificpcs[,i]=scale(pcs[,i])
  plot(seq(from=1901,length.out=nyrs), scale(pcs[,i]),type="b",xlab="Year",ylab=paste("PC",i),main=paste("Temporal Component (eigenvalue) number",i))
}
```

#### First four Principal Components (time)

```{r fig.width=16, fig.height=16}
par(mfrow=c(4,4))
par(pty="m")
for(i in 1:4){
  for(j in 1:4){
    fit=locfit(westernpcs[,j]~pacificpcs[1:113,i],alpha=1,deg=1)
      plot(fit,get.data=TRUE,ylab=ifelse(j==1,paste("Precip",colnames(westernpcs)[i]),""),xlab=ifelse(i==4,paste("Pac SST",  colnames(pacificpcs)[j]),""))   
    # }
    }
  }
```
All relations show low correlation, indicating that the results may be wrong.Relation Pacific SST PC2 - PreciP PC1 has a significant parameter in a linear regression, but R2 is only 0.07.

#### Aux code to call libraries, own functions and variables
```{r sources, results=FALSE}
##
## Source Libraries
##
libr=c("magrittr","maps","akima","fields","tidyr","locfit")
options(warn=-1)
suppressPackageStartupMessages(lapply(libr, require,character.only = TRUE))


##
## Source data
##


##
## Source functions
## 

```