---
title: "HW#2e8-main"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, ref.label=c("sources1","sources2")}
knitr::opts_chunk$set(echo = TRUE)

```

## Joint Spatial Analysis & Field Forecasting using CCA

One of the objectives of multivariate analysis is to enable multivariate forecasting. Here we wish to predict the winter precipitation over Arizona as a function of winter Pacific SSTs. We use CCA for this and the simplified steps are described below.

### (i) Take the leading ~ 5 PCs of the winter SSTs and perform CCA with the leading ~ 5 PCs of winter precipitation.

```{r}
# load data

## Western US precipitation
winterprecip = matrix(scan("http://civil.colorado.edu/%7Ebalajir/CVEN6833/HWs/HW-2-2018/western-US-winterprecip-1901-2014.txt"),ncol=82,byrow=T)
winterprecip = winterprecip[,2:82] # years = 1901:2013

## Pacific SST
load("C:/Users/alexb/Google Drive/CVEN 6833 ADAT/zz Homeworks/HW#2/CVEN6833-HW-2/problem3.RData")

## PCA on Western US precip
p.svd = winterprecip %>% scale(.) %>% var(.) %>% svd(.) # Eigen decomposition
p.pcs= t(t(p.svd$u) %*% t(scale(winterprecip))) %>% scale(.) # Principal Components
p.lambdas=(p.svd$d/sum(p.svd$d)) # Eigenvalues

## PCA on Pacific SST
t.svd = sstannavg %>% scale(.) %>% var(.) %>% svd(.) # Eigen decomposition
t.pcs=t(t(t.svd$u) %*% t(scale(sstannavg))) %>% scale(.) # Principal Components
t.lambdas=(t.svd$d/sum(t.svd$d)) # Eigenvalues
```

DO CCA

```{r}
npc= 5 # N. of PCA to do CCA

```


### (ii) Fit a regression for each canonical variate of precipitation related to canonical variate of SSTs. 

#### Use these regressions to predict the flow variates and back transform them to the original space – i.e., the precipitation space.

### (iii) Evaluate the performance by computing R^2^ between the observed and predicted summer precipitation at each grid point.

### (iv) Use the model to predict the maximum precipitation for the last two years.




#### Aux code to source libraries, functions and variables

```{r sources1, results=FALSE}
##
## Source Libraries
##

libr=c("kohonen","ggplot2","fields","magrittr","ggpubr","scales")
options(warn=-1)
suppressPackageStartupMessages(lapply(libr, require,character.only = TRUE))

##
## Source data
##


##
## Source functions
## 

```
