---
title: "HW#2-main"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1. Bayesian Hierarchical Spatial Model
### (Do HW#1.11) Estimate the spatial surface of January precipitation with a Bayesian Hierarchical Spatial Model (see Verdin et al. 2015)

Initialize (source) libraries, functions:
```{r }
setwd("C:/Users/alexb/Google Drive/CVEN 6833 ADAT/zz Homeworks/HW#2/CVEN6833-HW-2")
source("source_file.R")
set.seed(1) # Set seed to repodruce results
```

### Fit a variogram on the residuals of a best GLM
The best GLM (HW#1-3) for January was the one with 7 covariates, Gamma family and log link function: 
Y ~ Lat + Long + Elev + LatLong + LatElev + LongElev + All.

```{r}
# Adding columns to January and fitting best GLM 
jan$LatLong=jan$Lat*jan$Long
jan$LatElev=jan$Lat*jan$Elev
jan$LongElev=jan$Long*jan$Elev
jan$All=jan$Lat*jan$Long*jan$Elev

# Fit best GLM
y=jan$Prec/10 #Use cm instead of mm
xx=jan[!colnames(jan) %in% "Prec"]
yglm=glm(y~.,data = xx,family = Gamma(link = "log"))

# Fit a variogram 
geod = as.geodata(cbind(xx$Long,xx$Lat,yglm$residuals))
geod = jitterDupCoords(geod,max=0.01) #jitter close values
vg = variog(geod,uvec = 12)
plot(vg)

```

With the results from the variogram we can assign the initial value of the Monte Carlo Markov Chain
```{r}
# Sigma, phi, (nugget set to zero)
sigma.sq = median(vg$v)
sigma.sq.lo = 0.25*sigma.sq
sigma.sq.hi = 1.75*sigma.sq
phi.val = median(vg$u) 
phi.lo = 0.25*phi.val
phi.hi = 1.75*phi.val

n.samples <- 1000 # Input number of samples
coords <- geod$coords # coordinates of observations

# prior distributions defined (limited distributions available... see help(spLM)
priors <- list("beta.flat", "phi.unif"=c(phi.lo,phi.hi),
               "sigma.sq.ig"=c(sigma.sq.lo,sigma.sq.hi))

# starting values for MCMC resampling
starting <- list("phi"=phi.val, "sigma.sq"=sigma.sq)#, "tau.sq"=100)

# adjustment factor for MCMC sampling routine
tuning <- list("phi"=1, "sigma.sq"=1)#, "tau.sq"=100) # WJR why set these to 1??
```

### Perform Monte Carlo Markov Chain Analysis and plot posterior values

```{r}
bat.fm <- spLM(formula=y~., data=xx, coords=coords, priors=priors, tuning=tuning, starting=starting, cov.model = "exponential", n.samples = n.samples, verbose = FALSE, n.report = 1000)
# burn-in samples
bat.fm <- spRecover(bat.fm, start=1, thin=1, verbose=F)

# plot posterior pdfs
beta = bat.fm$p.beta.recover.samples
theta = bat.fm$p.theta.recover.samples
par(mfrow=c(2,2))
plot(beta,auto.layout=FALSE)
plot(theta)
par(mfrow=c(1,1))
```
The results from the MCMC are used to predict values on the grid using a Krigging model on the residuals (This chunk is not evaluated in the knit version and the results are incorporated in the next chunk from a backup file containing the results calculated independtly).

```{r eval=FALSE}
n.samples2 = floor(n.samples*(2/3)) # consider only the last 2/3 of the MCMC simulations
dropped=n.samples-n.samples2
yHBM = matrix(nrow=length(dem[,1]), ncol=n.samples2) # store y values at each dem point for each sample
ySE = matrix(nrow=length(dem[,1]), ncol=n.samples2) # same for SE

xx=jan[!colnames(jan) %in% "Prec"] # ind.var for Krigging model

dem$LatLong=dem$Lat*dem$Long #creating DEM with 7 variables as xx
dem$LatElev=dem$Lat*dem$Elev
dem$LongElev=dem$Long*dem$Elev
dem$All=dem$Lat*dem$Long*dem$Elev

# Loop over Krigging at each sample point
for(i in 1:n.samples2){
  ykrig=Krig(xx,yglm$residuals,rho = theta[dropped+i,1],theta = theta[dropped+i,2],m = 1)    # Krig model
  y2=predict.Krig(ykrig,x=dem,drop.Z = TRUE) # Predict residual from Krig at DEM locations
  ySE[,i]=predictSE(ykrig,x=dem,drop.Z = TRUE)                 # Predict SE of Krig at DEM 
  
  y1=beta[i,1]+sum(dem[i,]*beta[,2:8])                              # Predict y from MCMC
  yHBM[,i]=y1+y2                                                  # Add MCMC and residuals
  if(i%%100==0){sprintf("Iteration %s complete",i)}
}
```
The results from krigging over the MCMC sample are used to plot the posterior parameters of the Bayesian Hierarchical Spatial Model: histogram of precipitation, spatial map of mean, and spatial map of standard error.

```{r}
load("C:/Users/alexb/Google Drive/CVEN 6833 ADAT/zz Homeworks/HW#2/CVEN6833-HW-2/backupv1.RData")
yHBM=backup1 #backup files
ySE=backup2
mean.y = apply(yHBM, 1, FUN = mean) # posterior mean of precipitation
mean.yse = apply(ySE, 1, FUN = mean) # posterior standard error

par(mfrow=c(1,1))
hist(yHBM, main = "Histogram of Precipitation Posterior")

#Create spatial data to plot
X_sp<-getSpatialDataFrame(cbind(xdem,mean.y))
X_sp<-NAD83(X_sp)
X_ppp<-as.ppp(data.frame(X_sp@coords,X_sp@data$`mean.y`), W=colf_owin)

zz_sp<-getSpatialDataFrame(cbind(xdem,mean.yse))
zz_sp<-NAD83(zz_sp)
zz_ppp<-as.ppp(data.frame(zz_sp@coords,zz_sp@data$`mean.yse`), W=colf_owin)

# plot calls
generateplot(X_ppp)
title(main ='January - BHSM - Predicted Precipitation of Precip. (cm)')

generateplot(zz_ppp)
title(main='January - BHSM - Predicted Standard Error of Precip. (cm)')
```

