---
title: "HW#2e5_main"
output: html_document
---

```{r setup, include=FALSE, ref.label="sources"}
knitr::opts_chunk$set(echo = TRUE)

# Source data, functions, librarires (end of document)
```

## Exercise 5. Classification And Regression Trees (CART) 

### (a) Pick a locations in Arizona and fit a CART model to its seasonal precipitation along with 3 PCs of the winter SSTs over Pacific

```{r fig.width=14,fig.height=6}
# Source data, functions, librarires are at the end of document (Western US precipitation and Arizona Point)

load("C:/Users/alexb/Google Drive/CVEN 6833 ADAT/zz Homeworks/HW#2/CVEN6833-HW-2/pacificSST.RData") # 4 PC from SST

precip_dt=cbind(winterprecip[,2],pacificpcs[1:113,1:3]) %>% as.data.frame(.) %>% setNames(.,c("Y","X1","X2","X3")) # data

# tree.precip = tree(Y ~ X1+X2+X3, data = precip_dt)
tree.precip2 = rpart(Y ~ X1+X2+X3, data=precip_dt, method="anova") # rpart package

summary(tree.precip2)

# plot(tree.precip) #unprunned tree
# text(tree.precip, all=TRUE, cex=.6)

plot(tree.precip2,main="Classification Tree for Arizona winter precipitation")
text(tree.precip2, all=TRUE, cex=.8)
```

The fit (or tree prune) is done selecting the tree with a balance between low cross-validated erro and number of splits. Selecting the tree with the complexity parameter associated with the smallest cross-validated error leads to the original tree.
```{r}
printcp(tree.precip2) # display the results 

plotcp(tree.precip2) # Plots the cross-validation results

treeFit=rpart::prune(tree.precip2, cp=tree.precip2$cptable[3,1]) 
#min cp is: cp= tree.precip2$cptable[which.min(tree.precip2$cptable[,"xerror"]),"CP"]

plot(treeFit,uniform=TRUE,main="Pruned Classification Tree for Arizona winter precipitation")
text(treeFit, use.n=TRUE, all=TRUE, cex=.8)

treePred <- predict(treeFit,type = "matrix")
treeResid <- resid(treeFit)
myRange <- range(treePred, precip_dt$Y)

plot(precip_dt$Y, treePred, xlim = myRange, ylim = myRange, xlab="True value", ylab="Prediction", main = paste("Pred vs True for prunned tree"))
abline(0,1, col = "red")
```

#### Perform a drop-10% cross validation and boxplot correlation and RMSE

```{r}
nsim = 50
rmseskill = 1:nsim
corskill=1:nsim
N=length(precip_dt$Y)
for(i in 1:nsim){
  drop=sample(c(1:N),round(N*0.1))
  keep=setdiff(1:N,drop)
  tree.keep = rpart(Y ~ X1+X2+X3, data=precip_dt[keep,], method="anova") # rpart package
  tree.drop = predict(tree.keep,newdata = precip_dt[drop,])
  rmseskill[i]=sqrt(mean((precip_dt[drop,"Y"]-tree.drop)^2))
  corskill[i]=cor(precip_dt[drop,"Y"],tree.drop)
}
par(mfrow=c(1,2))
boxplot(rmseskill,varwidth=TRUE,main="RMSE Skill")
boxplot(corskill,varwidth=TRUE, main="COR Skill")
```

The RMSE observed is large and the observed correlation is low, as in exercise 3.

### (b) Fit a random forest model and compare the performance with (a) above

```{r}
forest.precip = randomForest(Y ~ X1+X2+X3, data = precip_dt)
summary(forest.precip)
plot(forest.precip)
rforestpred = predict(forest.precip)

forestResid <- resid(forest.precip)
myRange <- range(rforestpred, precip_dt$Y)

plot(precip_dt$Y, rforestpred, xlim = myRange, ylim = myRange, xlab="True value", ylab="Prediction", main = paste("Pred vs True for prunned tree"))
abline(0,1, col = "red")

```
Results show greater variability, but the low correlation makes them still stay away from the diagonal.

```{r sources, results=FALSE}
##
## Source Libraries
##
libr=c("magrittr","mclust","tree","sp","maps","ggplot2","rpart","randomForest")
options(warn=-1)
suppressPackageStartupMessages(lapply(libr, require,character.only = TRUE))


##
## Source data
##

### Western US winter precipitation data
N = 113 # Time window
wuplocs = matrix(scan("http://civil.colorado.edu/%7Ebalajir/CVEN6833/HWs/HW-2-2018/WesternUS-coords.txt"), ncol=5,byrow=T)
nlocs = dim(wuplocs)[1]
xlats = wuplocs[,3]
xlongs = -wuplocs[,4]

winterprecip = matrix(scan("http://civil.colorado.edu/%7Ebalajir/CVEN6833/HWs/HW-2-2018/western-US-winterprecip-1901-2014.txt"),ncol=nlocs+1,byrow=T)
years = 1901:2013
winterprecip = winterprecip[,2:(nlocs+1)]		#first column is year

### Trim arizona 
z=maps::map('state', 'arizona', fill = FALSE,plot = FALSE)
zxy=cbind(z$x,z$y)
# zxy=zxy[complete.cases(zxy),] #map is not a close d polygon, just pick one

indint=point.in.polygon(xlongs, xlats, zxy[,1], zxy[,2])==1
indrel=point.in.polygon(xlongs, xlats, zxy[,1], zxy[,2])==2
arlocs=as.data.frame(cbind(xlongs[indint],xlats[indint])) # points inside Arizona
ardata=cbind(arlocs,t(winterprecip[,indint]))

arind=2 # Selected point
arp=c(xlongs[2],xlats[2]) # Selected coordinates

plot(xlongs,xlats,col="grey")
lines( zxy[,1], zxy[,2],col="red")
points(xlongs[indint],xlats[indint],col="green")
points(xlongs[indrel],xlats[indrel],col="blue")
points(xlongs[2],xlats[2],col="red",pch=20)  # Plot point

##
## Source functions
## 

```
