---
title: "HW#2_e4_main"
output: html_document
---

```{r setup, include=FALSE, ref.label="sources"}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 4. PCA + Multinomial Regression 

### (a) Perform a PCA on the attribute data of the safety ocutcome of body part - show the Eigen spectrum and the leading four Eigen vectors

*bpartbin* is the categorical variable vector
*Xpredictors* are the 20 most relevant variables to body part safety
This data is obtained from the source file included at the end of the exercise.

```{r}
zsvd = Xpredictors %>% scale(.) %>% var(.) %>% svd(.) # scale # get variance # Eigen decomposition
pcs=t(t(zsvd$u) %*% t(scale(Xpredictors))) #Principal Components...
lambdas=(zsvd$d/sum(zsvd$d)) #Eigen Values.. - fraction variance explained 

plot(1:20, lambdas[1:20], type="l", xlab="Modes", ylab="Frac. Var. explained",main="Eigen spectrum") #plot spectrum
points(1:20, lambdas[1:20], col="red")

par(las=2) # horizontal y - axis label
par(mar=c(4,12,4,4)) # increase y-axis margin.

for(i in 1:4){ # plot the four eigenvectors against each attribute
  barplot(zsvd$u[,i],main=paste("Eigenvector",i,"influence on each attribute"),names.arg = colnames(Xpredictors),horiz=TRUE,cex.names=0.8)
  indexmax=sort(abs(zsvd$u[,i]),decreasing = TRUE,index.return = TRUE) # sort eigen by largest values 
  print(sprintf("Most influential attributes eigenvector %s: %s",i,paste(colnames(Xpredictors)[indexmax$ix[1:4]],collapse = " / "))) # print the four most influential attributes in eigenvector.
  }
```

### (b) Fit a best multinomial regression with the principal components as predictors to model the categorical probability of injuries to the five category of body parts and compute the RPSS

```{r}
pcs = cbind(data.frame(bpartbin),data.frame(pcs)) # dataset
zz = multinom(pcs$bpartbin ~ ., data = pcs[,1:6]) # Multinomial log-linear regression
N = log(length(bpartbin)) # penalty function for stepAIC

zbest=stepAIC(zz,k=N) # Check if the selected model is the best. It turns to be the best

# Prediction skill via RPSS
ypred = predict(zz, type = "probs") # prediction model

p.real=rep(NA,5) # Real probabilities
for(i in 1:5){
  p.real[i]=table(bpartbin)[i]/length(bpartbin)
}

rps(bpartbin,ypred,baseline = p.real)$rpss
```

This score, RPSS, answers the question: What is the relative improvement of the probability forecast over body injury data in predicting the category that the observations fell into (p.real based on counts vs.  multinom. regression prediction).

### (c) Repeat the process for Injury Severity data

*spartbin* is the categorical variable vector for injury severity data. Only 4 levels were found in the csv file despite being described as 6 in Tixier et al. (2016)
*Xpredsev* are the 20 most relevant attributes to injury severity part.
This data is obtained from the source file included at the end of the exercise.

#### PCA on the attribute data

```{r}
zsvd = Xpredsev %>% scale(.) %>% var(.) %>% svd(.) # scale # get variance # Eigen decomposition
pcs=t(t(zsvd$u) %*% t(scale(Xpredsev))) #Principal Components...
lambdas=(zsvd$d/sum(zsvd$d)) #Eigen Values.. - fraction variance explained 

plot(1:20, lambdas[1:20], type="l", xlab="Modes", ylab="Frac. Var. explained",main="Eigen spectrum") #plot spectrum
points(1:20, lambdas[1:20], col="red")

par(las=2) # horizontal y - axis label
par(mar=c(4,12,4,4)) # increase y-axis margin.

for(i in 1:4){ # plot the four eigenvectors against each attribute
  barplot(zsvd$u[,i],main=paste("Eigenvector",i,"influence on each attribute"),names.arg = colnames(Xpredsev),horiz=TRUE,cex.names=0.8)
  indexmax=sort(abs(zsvd$u[,i]),decreasing = TRUE,index.return = TRUE) # sort eigen by largest values 
  print(sprintf("Most influential attributes eigenvector %s: %s",i,paste(colnames(Xpredsev)[indexmax$ix[1:4]],collapse = " / "))) # print the four most influential attributes in eigenvector.
  }
```

#### Best multinomial regression with the PCA as predictors to model the probability of injuries
This is done for  each of  the five category of body parts and later compute the RPSS

```{r}
pcs = cbind(data.frame(spartbin),data.frame(pcs)) # dataset
zz = multinom(pcs$spartbin ~ ., data = pcs[,2:6]) # Multinomial log-linear regression
N = log(length(spartbin)) # penalty function for stepAIC

zbest=stepAIC(zz,k=N) # Check if the selected model is the best. A constant value is the best model

# Prediction skill via RPSS
ypred = predict(zz, type = "probs") # prediction model

p.real=rep(NA,4) # Real probabilities
for(i in 1:4){
  p.real[i]=table(spartbin)[i]/length(spartbin)
}

rps(spartbin,ypred,baseline = p.real)$rpss
```

According to the definition of the RPSS the improvement given by the model is really low, as it happens to Tixier et al. (2016) due to imbalanced categories.

#### Aux code to call libraries, own functions and variables
```{r sources, results=FALSE}
##
## Source Libraries
##
libr=c("magrittr","nnet","MASS","verification")
options(warn=-1)
suppressPackageStartupMessages(lapply(libr, require,character.only = TRUE))


##
## Source data
##

### Safety data
to.retain=20 # imp var to retain
imp.vars.energy=as.character(read.csv("gbm.importance.50.energy.csv",header=TRUE)[1:to.retain,"var"])
imp.vars.inj.type=as.character(read.csv("gbm.importance.50.code.csv",header=TRUE)[1:to.retain,"var"])
imp.vars.body=as.character(read.csv("gbm.importance.50.body.csv",header=TRUE)[1:to.retain,"var"])
imp.vars.severity=c("crane","heavy.material.tool","stairs","unpowered.tool","exiting.transitioning","heavy.vehicle","uneven.surface","drill","ladder","object.at.height","improper.procedure.inattention","unpowered.transporter","machinery","improper.security.of.materials","working.at.height","no.improper.PPE","small.particle","steel.steel.sections","stripping","scaffold")

# Body data
data.body = read.csv("data.body.part.csv") 
index = data.body %>% {colnames(.) %in% imp.vars.body} %>% which(.)

Xpredictors = data.body[,index]

bpartbin = data.body[,1] %>% as.character(.) 
bpartbin = bpartbin %>% {. == "head"} %>% replace(bpartbin, . , 1) %>% replace(., .== "neck", 2) %>% replace(., .== "trunk" , 3) %>% replace(., .== "upper extremities" , 4) %>% replace(., .== "lower extremities" , 5) %>% as.numeric(.) # Categorical variables

#Injury severity data
data.severity=read.csv("data.severity.csv")
index = data.severity %>% {colnames(.) %in% imp.vars.severity} %>% which(.)

Xpredsev = data.severity[,index]

spartbin = data.severity[,1] %>% as.character(.)
sevname=unique(spartbin)
spartbin = spartbin %>% {. == sevname[1]} %>% replace(spartbin, . , 1) %>% replace(., .== sevname[2], 2) %>% replace(., .== sevname[3] , 3) %>% replace(., .== sevname[4] , 4) %>% as.numeric(.) # Categorical variables



##
## Source functions
## 

```